The hydra and the rooster

### 2020-08-03 latex/chapter-hydras.tex
correction on pdf file

\vspace{4pt}
\emph{From Module~\href{../src/html/hydras.Hydra.Hydra_Definitions.html\#Hvariant}{Hydra.Hydra\_Definitions}}
\index{Coq!Techniques!Type classes}

@@

\begin{exercise}
 Prove that, if there is an instance of \texttt{Hvariant Lt wf\_Lt $B$ $m$}, then there exists no infinite battle in  $B$.
\end{exercise}

\subsection{A first theorem of impossibility}

@@

@@

Let us follow the same pattern as in Sect.~\ref{omega-case}.
First, we define an injection from type \texttt{t} into \texttt{Hydra}.
We associate with any pair $(i,j)$ the hydra with $i$ branches of length $2$ and
$j$ branches of length $1$.

@@

\begin{exercise}

@@

For instance , the sequence $\langle 4,4,2 \rangle$ is strictly greater than 
$\langle 4,3,3,3,3,3,3,2,2,2 \rangle$.
\end{itemize}

@@

\subsubsection{Conclusion/Teaser}

>>>>

\vspace{4pt}
\emph{From Module~\href{../src/html/hydras.Hydra.Hydra_Definitions.html\#Hvariant}{Hydra.Hydra\_Definitions}}
\index{Coq!Techniques!Type classes}
\label{sect:hvariant-def}

@@

\begin{exercise}
 Prove that, if there is an instance of (\texttt{Hvariant Lt wf\_Lt $B$ $m$}), then there exists no infinite battle in  $B$.
\end{exercise}

\subsection{Two small proofs of impossibility}
\index{Maths!Proofs of impossibility}

When one wants to prove a termination theorem with the help of a variant, 
one has to consider first a well-founded set $(A,<)$, then a strictly decreasing measure on this set.  The following two lemmas show that if  the order structure $(A,<)$ is too simple, it is useless to look for a convenient measure, which simply no exists. Such kind of result is useful, because it saves you time and effort.

\subsubsection{No variant on Peano.lt}
\label{omega-case}

@@

\begin{figure}[htb]
\centering
\begin{tikzpicture}[very thick, scale=0.5]
\node (foot) at (2,0) {$\bullet$};
\node (N1) at (2,2) {$\bullet$};
\node (N2) at (2,4) {$\Smiley[2][green]$};
\draw (foot) -- (N1);
\draw (N1) to  (N2);
\end{tikzpicture}
\caption{\label{fig:h-omega-omega}}
 The hydra \texttt{big\_h}.
\end{figure}

@@

Let us follow the same pattern as in Sect.~\ref{omega-case}.
First, we define an injection from type \texttt{t} into \texttt{Hydra}, by
 associating to  any pair $(i,j)$ the hydra with $i$ branches of length $2$ and
$j$ branches of length $1$.

@@

\index{Exercises}
\begin{exercise}

@@

For instance, the following inequality holds:
\[\langle 4,3,3,3,3,3,3,2,2,2 \rangle\,<\,\langle 4,4,2 \rangle\]
\end{itemize}

@@

\subsubsection{Conclusion}

### 2020-08-03 latex/chapter-ordinals.tex
correction on pdf file

\chapter{Introduction to ordinal numbers}

@@

\begin{figure}[h]
  \centering
\fbox{\Large
  \begin{minipage}{1.0\linewidth}
  \begin{align*}
     &\textcolor{blue}{0},\,1,2,3,4,5,6,7,8,9,10,11,12,13,14,15,16,17,\ldots\\
&\textcolor{red}{\omega},\,\omega+1,\omega+2,\omega+3,\ldots\\
&\textcolor{red}{\omega\times 2},\,\omega\times 2+1,\ldots, \textcolor{red}{\omega\times 3},\,\omega\times 3+1,\ldots, \textcolor{red}{\omega\times 4},\ldots,\\
&\textcolor{red}{\omega^2},\ldots, \textcolor{red}{\omega^2\times 42},\ldots,\textcolor{red}{\omega^3},\ldots, \textcolor{red}{\omega^4},\omega^4+1,\ldots,\\
&\textcolor{red}{\omega^\omega},\ldots, \textcolor{red}{\omega^\omega+\omega^7\times 8},\ldots\textcolor{red}{\omega^\omega\times 2},\omega^\omega\times 2+1, \ldots,\\
&\textcolor{red}{\omega^{\omega^\omega}},\ldots, \textcolor{red}{\omega^{\omega^\omega}+\omega^\omega\times 42+ \omega^{55}+\omega}, \ldots, \textcolor{red}{\omega^{\omega^{\omega+1}}}, \omega^{\omega^{\omega+1}}+1,\dots\\
& \textcolor{red}{\epsilon_0 (= \omega^{\epsilon_0)}}, \epsilon_0+1, \epsilon_0+2, \epsilon_0+3, \ldots,\\
& \textcolor{red}{\epsilon_1}, \ldots, \textcolor{red}{\epsilon_2}, \ldots, \textcolor{red}{\epsilon_\omega},\ldots \\
& \textcolor{red}{\Gamma_0}, \Gamma_0+1, \Gamma_0+2, \Gamma_0+3,\ldots, \textcolor{red}{\Gamma_0+\omega}, \ldots,\\
&\ldots
  \end{align*}   
  \end{minipage}}

@@

Let us comment some features of this table.

\begin{itemize}
\item The ordinals are listed in a strictly increasing order. 
\item Dots : ``$\ldots$'' stand for an infinite sequence of ordinals, not shown for lack of space. For instance, the ordinal $42$ is not shown in the first line, but it exists, betaween $41$ and $43$.
\item Each ordinal in black is an immediate successor of anotter ordinal. We call it a \emph{successor} ordinal. For instance, $43$ is the successor of $42$, and $\omega^4+1$ the successor of $\omega^4$.
\item Ordinals that  follow dots immediately (written in red) are called \emph{limit ordinals}. With respect to the order induced by this sequence, any limit ordinal $\alpha$ is the least upper bound of  the set $\mathbb{O}_\alpha$ of all ordinals strictly less than $\alpha$.
\item
For instance $\omega$ is the least upper bound of the set of all finite ordinals (in the first line). $\omega$ is also the first limit ordinal, and the first infinite ordinal, in the sense that 
the set $\mathbb{O}_\omega$ is infinite.
\item The ordinal $\epsilon_0$ is the first number which is equal to its own exponential of base $\omega$. It plays an important role in proof theory, and is particularly studied in chapters~\ref{chap:T1} to \ref{chap:alpha-large}.
\item All the ordinals shown in this list are either the ordinal \textcolor{blue}{$0$},
a successor ordinal, or a \textcolor{red}{limit ordinal}.
\end{itemize}

@@

\begin{align*}
(i,j) < (i,k) & \;\textbf{if} \; j < k\\
(0,k) < (1,l) & \;\textbf{for\,any}\;k \;\textbf{and} \; l
\end{align*}

@@

Fortunately, the ordinals we need for  studying hydra battles are much simpler than Schütte's, and can be represented as rather simple data types in \gallina. So, we will use \emph{ordinal notations} (also called \emph{[ordinal] notation systems}). Let $\alpha$ be some countable ordinal; In \coq{} terms, an ordinal notation for $\alpha$ will be a data type for representing all ordinals below $\alpha$ (excluding $\alpha$) and a bunch of functions for comparing ordinals, doing some arithmetic, inspecting the topology of the order (limits, successors), and proofs of well-foundedness). These functions should be proved correct, with respect to Schütte's model.

Our definition of ordinal notation borrows much material from \coq's standard library.

@@

The library ~\href{https://coq.inria.fr/distrib/current/stdlib/Coq.Classes.RelationClasses.html}{%
Classes.RelationClasses} contains some definitions and facts about binary relations, among them strict orders.

@@

\begin{Coqanswer}
Inductive
le_AsB (A B : Type) (leA : A -> A -> Prop) (leB : B -> B -> Prop)
  : A + B -> A + B -> Prop :=
    le_aa : forall x y : A, leA x y -> le_AsB A B leA leB (inl x) (inl y)
  | le_ab : forall (x : A) (y : B), le_AsB A B leA leB (inl x) (inr y)
  | le_bb : forall x y : B, leB x y -> le_AsB A B leA leB (inr x) (inr y)
\end{Coqanswer}

@@

\vspace{4pt}
\noindent\emph{From Module~\href{../src/html/hydras.Prelude.ON_Omega_plus_omega.html}{Prelude.ON\_Omega\_plus\_omega}}

@@

\index{Coq!Techniques!Coercions} 
\begin{remark}
Beware of coercions and notation scopes!

@@

Using these definitions , we can prove the following lemmas:

@@

\noindent\emph{From Module~\href{../src/html/hydras.Prelude.ON_Omega_plus_omega.html}{Prelude.ON\_Omega\_plus\_omega}}

\begin{Coqsrc}
 Lemma omega_is_limit : Limit omega.

Lemma limit_is_omega alpha : Limit alpha -> alpha = omega.

@@

@@

The following class specifies that a notation \texttt{OA} describes a segment $[0,\alpha)$,
where $\alpha$ is a countable ordinal à la Schütte.

@@

For instance, the following theorem tells that \texttt{Epsilon0}, our notation system for the segment $[0,\epsilon0)$ is correct with respect to the countqable ordinal $\epsilon_0$
(see chapter~\ref{chap:schutte} for more details).

@@

\begin{Coqsrc}
Class  ON_Isomorphic 
       `(OA : @ON A ltA compareA)
       `(OB : @ON B ltB  compareB)
       (f : A -> B)
       (g : B -> A):=
  {
  iso_compare: forall x y : A,  compareB (f x) (f y) =
                                 compareA x y;
  iso_inv1 : forall a, g (f a)= a;
  iso_inv2 : forall b, f (g b) = b}.
\end{Coqsrc}

@@

\section*{Other ordinal notations}

>>>>

\chapter{Introduction to ordinal numbers and ordinal notations}

@@

\begin{figure}[h]
  \centering
\fbox{\Large
  \begin{minipage}{1.0\linewidth}
  \begin{align*}
     &\textcolor{blue}{0},\,1,2,3,4,5,6,7,8,9,10,11,12,13,14,15,16,17,\ldots\\
&\textcolor{red}{\omega},\,\omega+1,\omega+2,\omega+3,\ldots\\
&\textcolor{red}{\omega\times 2},\,\omega\times 2+1,\ldots, \textcolor{red}{\omega\times 3},\,\omega\times 3+1,\ldots, \textcolor{red}{\omega\times 4},\ldots,\\
&\textcolor{red}{\omega^2},\ldots, \textcolor{red}{\omega^2\times 42},\ldots,\textcolor{red}{\omega^3},\ldots, \textcolor{red}{\omega^4},\omega^4+1,\ldots,\\
&\textcolor{red}{\omega^\omega},\ldots, \textcolor{red}{\omega^\omega+\omega^7\times 8},\ldots,\textcolor{red}{\omega^\omega\times 2},\omega^\omega\times 2+1, \ldots,\\
&\textcolor{red}{\omega^{\omega^\omega}},\ldots, \textcolor{red}{\omega^{\omega^\omega}+\omega^\omega\times 42+ \omega^{55}+\omega}, \ldots, \textcolor{red}{\omega^{\omega^{\omega+1}}}, \omega^{\omega^{\omega+1}}+1,\dots\\
& \textcolor{red}{\epsilon_0 (= \omega^{\epsilon_0)}}, \epsilon_0+1, \epsilon_0+2, \epsilon_0+3, \ldots,\\
& \textcolor{red}{\epsilon_1}, \ldots, \textcolor{red}{\epsilon_2}, \ldots, \textcolor{red}{\epsilon_\omega},\ldots \\
& \textcolor{red}{\Gamma_0}, \Gamma_0+1, \Gamma_0+2, \Gamma_0+3,\ldots, \textcolor{red}{\Gamma_0+\omega}, \ldots,\\
&\ldots
  \end{align*}   
  \end{minipage}}

@@

Let us comment some features of this table:

\begin{itemize}
\item The ordinals are listed in a strictly increasing order. 
\item Dots : ``$\ldots$'' stand for an infinite sequence of ordinals, not shown for lack of space. For instance, the ordinal $42$ is not shown in the first line, but it exists, betaween $41$ and $43$.
\item Each ordinal in black is an immediate successor of anotter ordinal. We call it a \emph{successor} ordinal. For instance, $43$ is the successor of $42$, and $\omega^4+1$ the successor of $\omega^4$.
\item Ordinals that  follow dots immediately (displayed in red) are called \emph{limit ordinals}. With respect to the order induced by this sequence, any limit ordinal $\alpha$ is the least upper bound of  the set $\mathbb{O}_\alpha$ of all ordinals strictly less than $\alpha$.
\item
For instance $\omega$ is the least upper bound of the set of all finite ordinals (in the first line). $\omega$ is also the first limit ordinal, and the first infinite ordinal, in the sense that 
the set $\mathbb{O}_\omega$ is infinite.
\item The ordinal $\epsilon_0$ is the first number which is equal to its own exponential of base $\omega$. It plays an important role in proof theory, and is particularly studied in chapters~\ref{chap:T1} to \ref{chap:alpha-large}.
\item All the ordinals shown in this list are either the ordinal \textcolor{blue}{$0$},
a successor ordinal, or a \textcolor{red}{limit ordinal}.
\end{itemize}

@@

\begin{align*}
(i,j) < (i,k) & \;\textit{if} \; j < k\\
(0,k) < (1,l) & \;\textit{for\,any}\;k \;\textit{and} \; l
\end{align*}

@@

Fortunately, the ordinals we need for  studying hydra battles are much simpler than Schütte's, and can be represented as quite simple data types in \gallina. So, we will use \emph{ordinal notations} (also called \emph{[ordinal] notation systems}).

Let $\alpha$ be some (countable) ordinal; 
in \coq{} terms, we call ordinal notation for $\alpha$ a structure composed 
of:
\begin{itemize}
\item A data type $A$ for representing all ordinals strictly below $\alpha$,
\item A well founded order $<$ on $A$, 
\item A Function for comparing two ordinals, which implies that the reflexive closure of $<$ is a \emph{total order}.
\end{itemize}

Such a structure can be proved correct relatively to another ordinal notation or
to Schütte's model.

@@

The library ~\href{https://coq.inria.fr/distrib/current/stdlib/Coq.Classes.RelationClasses.html}{%
Coq.Classes.RelationClasses} contains some definitions and facts about binary relations, among them strict orders.

@@

\begin{Coqanswer}
Inductive
le_AsB (A B : Type) (leA : A -> A -> Prop) (leB : B -> B -> Prop)
  : A + B -> A + B -> Prop :=
| le_aa : forall x y : A, leA x y -> le_AsB A B leA leB (inl x) (inl y)
| le_ab : forall (x : A) (y : B), le_AsB A B leA leB (inl x) (inr y)
| le_bb : forall x y : B, leB x y -> le_AsB A B leA leB (inr x) (inr y)
\end{Coqanswer}

@@

\vspace{4pt}
\noindent\emph{From Module~\href{../src/html/hydras.OrdinalNotations.ON_Omega_plus_omega.html}{OrdinalNotations.ON\_Omega\_plus\_omega}}

@@

\label{warning:coercions}
\index{Coq!Techniques!Coercions} 
\begin{remark}
Beware of coercions and notation scopes!

@@

Using these definitions , we can prove the following lemma:

@@

\noindent\emph{From Module~\href{../src/html/hydras.OrdinalNotations.ON_Omega_plus_omega.html}{OrdinalNotations.ON\_Omega\_plus\_omega}}

\begin{Coqsrc}
Lemma limit_iff (alpha : t) : Limit alpha <-> alpha = omega.
\end{Coqsrc}

@@

\index{Projects}
\begin{project}
Adapt the definition of \texttt{Hvariant} (Sect.~\ref{sect:hvariant-def}) in order to
have an ordinal notation as argument. Prove that if $O_A$ is a sub-notation of $O_B$, then any variant defined on  $O_A$ can be automatically transformed into 
a variant on $O_B$.
\end{project}

@@

The following class specifies that a notation \texttt{OA} describes a segment $[0,\alpha)$,
where $\alpha$ is a countable ordinal \emph{à la}  Schütte.

@@

For instance, the following theorem tells that \texttt{Epsilon0}, our notation system for the segment $[0,\epsilon0)$ is correct with respect to the countable ordinal $\epsilon_0$
(see chapter~\ref{chap:schutte} for more details).

@@

\begin{Coqsrc}
Class  ON_Iso 
       `(OA : @ON A ltA compareA)
       `(OB : @ON B ltB  compareB)
       (f : A -> B)
       (g : B -> A):=
  {
  iso_compare: forall x y : A,  compareB (f x) (f y) =
                                 compareA x y;
  iso_inv1 : forall a, g (f a)= a;
  iso_inv2 : forall b, f (g b) = b}.
\end{Coqsrc}

@@

\section{Other ordinal notations}

### 2020-08-03 latex/chapter-schutte.tex
correction on pdf file

Our library 
\href{../src/html/hydras.Schutte.Injection_from_T1.html}%
{\texttt{Schutte.Injection\_from\_T1.v}} establishes the link between two very different modelizations of ordinal numbers. In other words, it ``validates'' a data structure in terms of
a classical mathematical discourse considered as a model. 
First, we define a function from \texttt{T1} into  \texttt{Ord} by structural recursion.

>>>>

Our library 
\href{../src/html/hydras.Schutte.correctness_E0.html}%
{\texttt{Schutte.Correctness\_E0}} establishes the link between two very different modelizations of ordinal numbers. In other words, it ``validates'' a data structure in terms of
a classical mathematical discourse considered as a model. 
First, we define a function from \texttt{T1} into  \texttt{Ord} by structural recursion.

### 2020-08-03 latex/chapter-termination.tex
correction on pdf file

\index{Types!T2@T2 (ordinals in Veblen normal form)}
\emph{Please look also at sub-directory
href{../src/html/hydras.Gamma0.T2.html}{Gamma0.T2.html}}.

@@

Another useful ordinal notation could be used for denoting the ordinals strictly less than $\omega^\omega$, but \texttt{list nat} could be used as well.

\begin{Coqsrc}
Inductive T0 : Set := 
  zero : T0
| cons : nat -> T0 -> T0.
\end{Coqsrc}

@@

\paragraph{Remark}
For simplicity's sake, we chosed to forbid  expressions of the form $\omega^\alpha\times 0 + \beta$. Thus, the contruction (\texttt{ocons $\alpha$ $n$ $\beta$} is intented to represent the
ordinal $\omega^\alpha\times(n+1)+\beta$ and not $\omega^\alpha\times n+\beta$.
In a future version, we should replace  the type \texttt{nat} with \texttt{positive} in \texttt{T1}'s 
definition. But this replacement would take a lot of time \dots.

@@

\index{Coq!Techniques!Coercions}
\index{Functions!Coercions@Coercions (from nat to ordinal types)}
\begin{remark}
\label{warning:coercions}
The use of coercions like \texttt{fin} allow us to be close to the mathematical tradition where natural numbers are ordinals too.
Nevertheless, it may happen that a goal like \texttt{3 < 5} could be 
interpreted as \texttt{(lt (fin 3) (fin 5))},  depending on the current notation scope.  
When this misinterpretation happens, tactics like \texttt{auto with arith}, \texttt{lia} do not work!
Thus, it is useful to write \texttt{(3 < 5)\%nat}  an inequality between two natural numbers. 
\end{remark}

@@

\index{Notations!omega@omega (the least infinite ordinal)}
\begin{Coqsrc}
Notation omega := (ocons (ocons zero 0 zero) 0 zero).
\end{Coqsrc}

@@

\begin{remark}
\label{sec:orgheadline69}
The name \(\varphi_0\)
   comes from ordinal numbers theory. In~\cite{schutte}, Schütte defines 
$\varphi_0$  as the ordering (\emph{i.e.} enumerating) function of the set  of \emph{additive principal ordinals} \emph{i.e.} strictly positive ordinals $\alpha$ that verify $\forall \beta<\alpha, \beta+\alpha=\alpha$. For Schütte,  $\omega^\alpha$ is just a notation for $\varphi_0(\alpha)$.  See also Chapter~\vref{chap:schutte}.
\end{remark}

@@

The ordinal $\epsilon_0$, although not represented by a finite term in Cantor normal form, is approximed by the sequance of $\omega$-towers.

@@

\begin{Coqsrc}
Theorem lt_irrefl : forall alpha, ~ lt alpha  alpha.

Theorem lt_trans :
forall alpha beta: T1,
 lt alpha  beta -> 
 forall gamma, lt beta gamma -> lt alpha gamma.

@@

\subsubsection{A Predicate for characterizing normal forms}
\label{sec:orgheadline74}
\label{sec:orgheadline75}
We note that our data-type \texttt{T1} allows us to write expressions that
are not properly in Cantor normal form as specified in Section \ref{sec:epsilon0-intro}.
For instance, consider the following term of type  \texttt{T1}.

@@

Definition nf alpha :Prop := nf_b alpha.
\end{Coqsrc}

@@

\emph{From Module~\href{../src/html/hydras.Schutte.Injection_from_T1.html}{Schutte.Injection\_from\_T1}}

>>>>

\index{Types!T2@T2 (ordinals in Veblen normal form)}
\emph{Please look also at the library
\href{../src/html/hydras.Gamma0.T2.html}{Gamma0.T2.html}}.

@@

@@

\paragraph{Remark}
For simplicity's sake, we chosed to forbid  expressions of the form $\omega^\alpha\times 0 + \beta$. Thus, the contruction (\texttt{ocons $\alpha$ $n$ $\beta$} is intented to represent the
ordinal $\omega^\alpha\times(n+1)+\beta$ and not $\omega^\alpha\times n+\beta$.
In a future version, we should replace  the type \texttt{nat} with \texttt{positive} in \texttt{T1}'s 
definition. But this replacement would take a lot of time \dots{}

@@

\index{Coq!Techniques!Coercions}
\index{Functions!Coercions@Coercions (from nat to ordinal types)}
\begin{remark}
Please refer to the remark~\pageref{waening:coercions} about the use of coercions.
% The use of coercions like \texttt{fin} allow us to be close to the mathematical tradition where natural numbers are ordinals too.
% Nevertheless, it may happen that a goal like \texttt{3 < 5} could be 
% interpreted as \texttt{(lt (fin 3) (fin 5))},  depending on the current notation scope.  
% When this misinterpretation happens, tactics like \texttt{auto with arith}, \texttt{lia} do not work!
% Thus, it is useful to write \texttt{(3 < 5)\%nat}  an inequality between two natural numbers. 
\end{remark}

@@

\index{Notations!omega@omega (the least infinite ordinal)}
\begin{Coqsrc}
Notation omega := (ocons (ocons zero 0 zero) 0 zero): t1_scope.
\end{Coqsrc}

@@

\begin{remark}
\label{sec:orgheadline69}
The name \(\varphi_0\)
   comes from ordinal numbers theory. In~\cite{schutte}, Schütte defines 
$\varphi_0$  as the ordering (\emph{i.e.} enumerating) function of the set  of \emph{additive principal ordinals} \emph{i.e.} strictly positive ordinals $\alpha$ that verify $\forall \beta<\alpha, \beta+\alpha=\alpha$. For Schütte,  $\omega^\alpha$ is just a notation for $\varphi_0(\alpha)$.  See also Chapter~\ref{chap:schutte} of this document.
\end{remark}

@@

The ordinal $\epsilon_0$, although not represented by a finite term in Cantor normal form, is approximed by the sequence of $\omega$-towers.

@@

\begin{Coqsrc}
Theorem lt_irrefl  alpha :  ~ lt alpha  alpha.

Theorem lt_trans (alpha beta gamma : T1) :
  lt alpha  beta -> lt beta gamma -> lt alpha gamma.

@@

\subsubsection{A predicate for characterizing normal forms}
\label{sec:orgheadline74}
\label{sec:orgheadline75}
We note that our data-type \texttt{T1} allows us to write expressions that
are not properly in Cantor normal form as specified in Section \ref{sec:epsilon0-intro}.
For instance, consider the following term of type  \texttt{T1}.

@@

Definition nf alpha: Prop := nf_b alpha.
\end{Coqsrc}

@@

\emph{From Module~\href{../src/html/hydras.Schutte.Schutte.Correctness_E0.html}{Schutte.Schutte.Correctness\_E0}}
